Fire Safety Regulation Violations in Toronto Residential Properties: Analysis by Property Type and Inspection Duration

### 2024-12-03 paper/paper.qmd
Updated addtional model details in paper

---
title: "Fire Safety Regulation Violations in Toronto Residential Properties: Analysis by Property Type and Inspection Duration"
subtitle: "Most properties would have a high probability of violation, whereeas high-rise buildings, rooming houses, and group homes have the highest risk"
author: 
  - Yunkai Gu
thanks: "Code and data are available at: [https://github.com/Kylie309/Toronto-Fire-Inspection](https://github.com/Kylie309/Toronto-Fire-Inspection)."
date: today
date-format: long
abstract: "This paper analyzes fire safety regulation violations in residential properties across Toronto, predicting the likelihood of violations based on the inspection duration and property type. By using a Bayesian logistic regression model, the study finds that property type plays a significant role in predicting violations, with high-rise buildings, rooming houses, and group homes at higher risk, and predicts with high confidence that most properties in the city would have a high probability of being observed to have a violation of fire safety regulations. These findings are important for improving fire safety protocols, as they suggest that targeting inspections towards high-risk property types could lead to more effective resource allocation and better safety outcomes. Findings in the study would motivate targeted inspections and interventions concentrated on properties that present the greatest risk, enhancing the effectiveness of fire safety efforts and helping to reduce potential fire accidents more efficiently."
format: pdf
number-sections: true
toc: true
prefer-html: true
bibliography: references.bib
---

@@

# Workspace setup
library(tidyverse)
library(janitor)
library(lubridate)
library(broom)
library(modelsummary)
library(rstanarm)
library(splines)
library(ggplot2)
library(arrow)
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(broom.mixed)

# Read the cleaned data
cleaned_data <- read_csv(
  file = here::here("data/02-analysis_data/fire_cleaned_data.csv")
  )
data <- read_csv(
  file = here::here("data/02-analysis_data/fire_cleaned_data.csv")
  )

@@

More detailed data cleaning process is included in [Appendix -@sec-appx-data].

The analyses presented in this paper were conducted using R programming language [@citeR]. The `tidyverse` packages [@citetidyverse] were used in the process of data simulation, testing beforehand. After the original raw data was downloaded by using `tidyverse` package [@citetidyverse], data cleaning process was done by using `tidyverse` package [@citetidyverse], `lubridate` package [@citelubridate], and `arrow` package [@citearrow]. We use `testthat` package [@citetest] to develop the test for structure and format of simulation and analysis data.

Then, models were constructed using `tidyverse` package [@citetidyverse], `lubridate` package [@citelubridate] and `rstanarm` [@citerstanarm] package. The model results were then presented by both the tables and graphs. Tables were made with `modelsummary` [@citemodelsummary] package, `broom.mixed` [@citebroom] package, `dplyr` [@citedplyr] package and `knitr` package. Graphs were made with `ggplot2` package [@citeggplot] and `RColorBrewer` package [@citercolorbrewer].

@@

# Create bar plot
data |>
  ggplot(aes(x = as.factor(violation), fill = as.factor(violation))) +
  geom_bar(color = "black") +
  labs(
    x = "Violation Status",
    y = "Count",
    fill = "Violation Status" # Legend title
  ) +
  scale_fill_manual(
    values = c("skyblue", "orange"),
    labels = c("No Violation Observed (Denoted as 0)", "Violation Observed (Denoted as 1)")
    ) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

@@

data |> 
  ggplot(aes(x = date_num)) + 
  geom_histogram(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(x = "Inspection Time Interval (Days)",
       y = "Counts of Properties") +
  theme_minimal()

@@

data_rise <- data %>% 
  filter(property_type == "High Rise" | property_type == "Low Rise")

data_other <- data %>% 
  filter(property_type != "High Rise" & property_type != "Low Rise")

@@

# Create bar plot for other property types
data_other |>
  ggplot(
    aes(x = as.factor(property_type), 
        fill = as.factor(property_type))
    ) +
  geom_bar(color = "black") +
  labs(
    x = "Property Type",
    y = "Count",
    fill = "Property Type" 
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

@@

data |> 
  ggplot(aes(x = as.factor(violation), fill = as.factor(violation))) +
  geom_bar(color = "black") +
  facet_wrap(~ property_type, scales = "free_y") + # Create one subplot per property type
  labs(
    x = "Violation Status",
    y = "Count",
    fill = "Violation Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_blank(), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 8), 
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.spacing = unit(1, "lines")
  ) +
  scale_fill_manual(values = c("skyblue", "orange"))
```

@@

# Use `tidy()` for a clean tabular format
library(broom.mixed)
tidy_stats <- tidy(model, conf.int = TRUE)

# Round all numeric columns to 3 decimal places
library(dplyr)
tidy_stats <- tidy_stats %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

# Create a table of parameter estimates
library(knitr)

tidy_stats %>%
  kable(format = "simple", col.names = c("Parameter", "Estimate", "Std.Error", "Conf.Low", "Conf.High"))

@@

# Additional data details {#sec-appx-data}

# Model details {#sec-model-details}

>>>>

---
title: "Fire Safety Regulation Violations in Toronto Residential Properties: Analysis by Property Type and Inspection Duration"
subtitle: "High Probability of Violations Across Most Properties, with High-Rise Buildings, Rooming Houses, and Group Homes Exhibiting the Highest Risk"
author: 
  - Yunkai Gu
thanks: "Code and data are available at: [https://github.com/Kylie309/Toronto-Fire-Inspection](https://github.com/Kylie309/Toronto-Fire-Inspection)."
date: today
date-format: long
abstract: "This paper analyzes fire safety regulation violations in residential properties across Toronto, predicting the likelihood of violations based on the inspection duration and property type, using data from fire safety inspections ended during 2024. By using a Bayesian logistic regression model, the study finds that property type plays a significant role in predicting violations, with high-rise buildings, rooming houses, and group homes at higher risk, and predicts with high confidence that most properties in the city would have a high probability of being observed to have a violation of fire safety regulations. These findings are important for improving fire safety protocols, as they suggest that targeting inspections towards high-risk property types could lead to more effective resource allocation and better safety outcomes. Findings in the study would motivate targeted inspections and interventions concentrated on properties that present the greatest risk, enhancing the effectiveness of fire safety efforts and helping to reduce potential fire accidents more efficiently."
format: pdf
number-sections: true
toc: true
prefer-html: true
bibliography: references.bib
---

@@

# Workspace setup
library(tidyverse)
library(janitor)
library(lubridate)
library(modelsummary)
library(rstanarm)
library(ggplot2)
library(arrow)
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(broom)
library(broom.mixed)

# Read the cleaned data
cleaned_data <- read_parquet(
  file = here::here("data/02-analysis_data/fire_cleaned_data.parquet")
  )
```

@@

@tbl-overview presents several sample lines of the cleaned dataset. More detailed data cleaning process is included in [Appendix -@sec-appx-data].

```{r}
#| echo: false
#| eval: true
#| label: tbl-overview
#| tbl-cap: "Preview of the Cleaned Dataset"
#| warning: false

cleaned_data |>
  head() |>
  kable(format = "simple")
```

The analyses presented in this paper were conducted using R programming language [@citeR]. The `tidyverse` packages [@citetidyverse] were used in the process of data simulation, testing and downloading. After the original raw data was successfully downloaded, data cleaning process was done by using `tidyverse` package [@citetidyverse], `lubridate` package [@citelubridate], `janitor` package [@citejanitor] and `dplyr` package [@citedplyr]. The analysis dataset after cleaning is saved as a parquet file using `arrow` package [@citearrow].

Then, models were constructed using `tidyverse` package [@citetidyverse], `lubridate` package [@citelubridate] and `rstanarm` [@citerstanarm] package. The model results were then presented by both the tables and graphs. Tables were made with `modelsummary` [@citemodelsummary] package, `broom.mixed` package [@citebroom], `dplyr` package [@citedplyr], `knitr` package [@citeknitr] and `kableExtra` package [@citekableextra]. Graphs were made with `ggplot2` package [@citeggplot].

@@

# Create bar plot
cleaned_data |>
  ggplot(aes(x = as.factor(violation), fill = as.factor(violation))) +
  geom_bar(color = "black") +
  labs(
    x = "Violation Status",
    y = "Count",
    fill = "Violation Status" # Legend title
  ) +
  scale_fill_manual(
    values = c("skyblue", "orange"),
    labels = c("No Violation Observed (Denoted as 0)", "Violation Observed (Denoted as 1)")
    ) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

@@

cleaned_data |> 
  ggplot(aes(x = date_num)) + 
  geom_histogram(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(x = "Inspection Time Interval (Days)",
       y = "Counts of Properties") +
  theme_minimal()
```

@@

data_rise <- cleaned_data %>% 
  filter(property_type == "High Rise" | property_type == "Low Rise")

data_other <- cleaned_data %>% 
  filter(property_type != "High Rise" & property_type != "Low Rise")

@@

# Create bar plot for other property types
data_other |>
  ggplot(
    aes(x = as.factor(property_type), 
        fill = as.factor(property_type))
    ) +
  geom_bar(color = "black") +
  labs(
    x = "Property Type",
    y = "Count",
    fill = "Property Type" 
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

@@

cleaned_data |> 
  ggplot(aes(x = as.factor(violation), fill = as.factor(violation))) +
  geom_bar(color = "black") +
  facet_wrap(~ property_type, scales = "free_y") + # Create one subplot per property type
  labs(
    x = "Violation Status",
    y = "Count",
    fill = "Violation Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_blank(), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 8), 
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.spacing = unit(1, "lines")
  ) +
  scale_fill_manual(values = c("skyblue", "orange"))
```

@@

# Use `tidy()` for a clean tabular format
tidy_stats <- tidy(model, conf.int = TRUE)

# Round all numeric columns to 3 decimal places
tidy_stats <- tidy_stats %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

# Create a table of parameter estimates
tidy_stats %>%
  kable(format = "simple", col.names = c("Parameter", "Estimate", "Std.Error", "Conf.Low", "Conf.High"))

@@

# Additional Data Details {#sec-appx-data}

The dataset cleaning process for preparation follows several steps:

1. The column names are standardized to snake_case using `janitor` package for consistency and readability.

2. The date information in the 'inspections_closeddate' and  'inspection_opendate' columns are initially recorded in inconsistent format. The date information is processed and stored as a valid date type in the columns, either from the datetime (ymd_hms) or date-only (ymd) parsing using `lubridate` package.

3. The dataset is then filtered to keep only cases (rows) where inspection processes were closed during 2024, i.e. 'inspections_closeddate' is on or after January 1, 2024.

4. Time interval between the start date and end date of the inspection (i.e. difference in days between 'inspections_closeddate' and 'inspections_opendate') is calculated and stored as a new numeric column called 'date_num'.

5. A new column named 'violation' is created to indicate whether there was a violation being detected in the inspection case. If the 'violations_item_number' presents value of 0, 'violation' is set to 0; otherwise, it is set to 1.

6. Finally, only necessary column are selected and retained in the cleaned dataset: 'property_type', 'inspections_opendate', 'inspections_closeddate', 'date_num', and 'violation'.

# Model Details {#sec-model-details}
